Dependently Typed Multi-stage Calculus

### 2019-11-26 slides/main.tex
Add future works.
Reflect Igarashi's comments.

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for wrong length vectors.
    \begin{verbatim}
 val vadd3 : (Vec -> Vec -> Vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : Vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{verbatim}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Dependent Types}
    \renewcommand{\V}{\text{Vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevents illegal use of {\verb|vadd3|}.
    \end{itemize}
 \begin{verbatim}
 - : [1;2]: Vec 2
 - : vadd3 : (Vec 3 -> Vec 3 -> Vec 3) code

@@

\begin{frame}[fragile]{\LTP [Hanada\&Igarashi'14]}
    Multi-stage calculus with Cross Stage Persistence (CSP)
    \begin{block}{Terms}
        \begin{tabbing}
        \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
        \> \( \mid \TB M \) \> (quote) \\
        \> \( \mid \TBL M \) \> (unquote) \\
        \> \( \mid \textbf{run}\ M \) \> (run) \\
        \> \( \mid \% M \) \> (CSP) \\
        \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \) \\
    \end{tabbing}
    \end{block}
\end{frame}

@@

    Want to use \( \text{mul5} \) in a code value.
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    CSP symbol \( \% \) {\bf{lift up}} an normal value into code value.
\end{frame}

@@

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Append function dependent types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (function dependent type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{Vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    % Note: We need CSP to use normal variables in code types.
    % \begin{itemize}
    %     \item OK  \( \Pi n: \TW (\text{Vec}\ \%n) \)
    %     \item NG  \( \Pi n: \TW (\text{Vec}\ n) \)
    % \end{itemize}
    \note{
    }
\end{frame}

@@

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    Introduce \QPercent.
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        }
    \end{center}
    Erase the symbol of \( \% \) under the condition.
    \begin{itemize}
        \item \( \text{Vec}\ \%3 \E \text{Vec}\ 3 \)
        \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
            \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
    \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Related Work}
    \begin{itemize}
        \item \LTP [Hanada\&Igarashi'14]
            \begin{itemize}
                \item Multi-stage programming including explicit CSP
            \end{itemize}
        \item Concoqtion [Forgarty et al.]
            \begin{itemize}
                \item Introduce indexed type into MetaOCaml.
                \item Types are indexed by Coq terms.
            \end{itemize}
    \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Define typing, kinding, well-formed kinding rules for multi-stage calculus.
        \item Define equality rules.
            \begin{itemize}
                \item Especially, dealing with CSP.
            \end{itemize}
        \item Prove type soundness.
    \end{itemize}
    \note{
    }
\end{frame}

>>>>

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{verbatim}
 val vadd3 : (Vec -> Vec -> Vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : Vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{verbatim}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Dependent Types}
    \renewcommand{\V}{\text{Vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevent illegal use of {\verb|vadd3|}.
    \end{itemize}
 \begin{verbatim}
 [1;2]: Vec 2
 vadd3 : (Vec 3 -> Vec 3 -> Vec 3) code

@@

\begin{frame}[fragile]{\LTP [Hanada\&Igarashi'14]}
    Multi-stage calculus with Cross Stage Persistence (CSP)
    \begin{block}{Terms}
        \begin{tabbing}
            \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
            \> \( \mid \TB M \) \> (quote) \\
            \> \( \mid \TBL M \) \> (unquote) \\
            \> \( \mid \textbf{run}\ M \) \> (run) \\
            \> \( \mid \% M \) \> (Cross Stage Persistence (CSP)) \\
            \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \) \\
        \end{tabbing}
    \end{block}
\end{frame}

@@

    Want to use \( \text{mul5} \) in a code value.
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    CSP symbol \( \% \) {\bf{lift up}} mul5 from stage 0 to stage 1.
\end{frame}

@@

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Append function dependent types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (function dependent type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{Vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    % Note: We need CSP to use normal variables in code types.
    % \begin{itemize}
    %     \item OK  \( \Pi n: \TW (\text{Vec}\ \%n) \)
    %     \item NG  \( \Pi n: \TW (\text{Vec}\ n) \)
    % \end{itemize}
    \note{
    }
\end{frame}

@@

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    Erase the CSP symbol \( \% \) of an closed term.
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        }
    \end{center}
    \begin{itemize}
        \item \( \text{Vec}\ \%3 \E \text{Vec}\ 3 \)
        \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
            \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
    \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Related Work}
    \begin{itemize}
        \item \LTP [Hanada\&Igarashi'14]
            \begin{itemize}
                \item Multi-stage programming including explicit CSP
            \end{itemize}
        \item Concoqtion [Forgarty et al.'14]
            \begin{itemize}
                \item Introduce indexed type into MetaOCaml.
                \item Types are indexed by Coq terms.
            \end{itemize}
    \end{itemize}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Define typing, kinding, well-formed kinding rules for multi-stage calculus.
        \item Define equality rules.
            \begin{itemize}
                \item Especially, dealing with CSP.
            \end{itemize}
        \item Prove type soundness.
        \item Future work: algorithmic typing
    \end{itemize}
    \note{
    }
\end{frame}
